---
title: "Working with Controlled Terminology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{controlled-terminology}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, message=FALSE}
library(metacore)
library(dplyr, warn.conflicts = FALSE)
library(tibble)
```

Controlled terminology is a cornerstone of clinical data
standardization. In the context of CDISC standards (SDTM and ADaM),
controlled terminology refers to predefined sets of valid values that
variables can take. These standardized values ensure consistency,
traceability, and compliance across clinical trial datasets.

#### Types of Controlled Terminology

-   **Codelist**: A mapping between codes and their human-readable
    decodes (e.g., "Y" = "Yes", "N" = "No")

-   **Permitted Values**: A simple list of allowed values with no decode
    mapping

-   **External Library References**: References to external controlled
    vocabulary

## Getting started with `get_control_term()`

The `get_control_term()` function is the primary tool for retrieving
controlled terminology metadata from a metacore object. Let's start with
the basics.

#### Loading your metadata

First, load your metadata specification into metacore. This example uses
the included mock specification:

```{r eval=TRUE}
# Load specification from Excel
metacore <- spec_to_metacore(
  metacore_example("p21_mock.xlsx"),
  verbose = "silent"
)

# Examine the structure
metacore
```

#### Simple case: Retrieving a Single Codelist

When a variable has a unique codelist across all datasets, retrieval is
straightforward:

```{r eval=TRUE}
# Get the codelist for AESEV (Adverse Event Severity)
# This variable has a single definition across datasets
severity_codes <- get_control_term(metacore, "AESEV")
severity_codes
```

Notice that `get_control_term()` returns a tibble
with `code` and `decode` columns. You can use this directly in your data
validation or mapping routines:

```{r eval=TRUE}
# Example: Create a mapping for recoding
severity_mapping <- severity_codes %>%
  tibble::deframe() # Convert to named vector for dplyr::recode()

print(severity_mapping)
```

#### Specifying a Dataset

When a variable appears across multiple datasets with the same codelist,
or when you want to be explicit about which dataset's definition you're
using, either subset your metacore object for a single dataset or
specify the `dataset` parameter:

```{r eval=TRUE}
# Option 1: Subset the specification object (recommended workflow)
suppae_spec <- select_dataset(metacore, "SUPPAE", verbose = "silent")
qval_codes1 <- get_control_term(suppae_spec, "QVAL")

# Option 2: Specify the dataset within `get_control_term`
qval_codes2 <- get_control_term(metacore, "QVAL", dataset = "SUPPAE")

# These operations return the same result
identical(qval_codes1, qval_codes2) # TRUE
```

> **Tip:** Subsetting your metacore object using `select_dataset()` is the
> recommended way to introduce metadata driven programming with metacore
> into your workflow. When developing clinical datasets, programmers tend
> to work on one dataset at a time. Subsetting your metadata ensures that
> you are always working with the right metadata for the job.

## Value-Level Metadata (VLM)

Value-level metadata (VLM) is an advanced feature that allows a single
variable to have different controlled terminology depending on the value
of another variable. This is common in CDISC datasets where different
parameter types may require different codelists.

#### **Understanding VLM Structure**

In standard (non-VLM) datasets, a variable like `AVALCAT1` (Analysis
Value Category) has a single codelist. However, with VLM, the same
variable might have multiple codelists depending on the parameter code
or other conditions.

For example:

-   `AVALCAT1` for `PARAMCD = "ADURD"` (Duration) might have codes: "1",
    "2", "3"

-   `AVALCAT1` for `PARAMCD = "ADURF"` (Frequency) might have codes:
    "A", "B", "C"

#### **Retrieving VLM Codelists**

When a variable has value-level metadata, you must specify which
condition's codelist you want using the `where` parameter:

```{r eval=TRUE}
# This example uses a VLM test specification
vlm_meta <- spec_to_metacore(
  metacore_example("vlm_test_spec.xlsx"),
  where_sep_sheet = FALSE,
  verbose = "silent"
)

adex_spec <- select_dataset(vlm_meta, "ADEX", verbose = "silent")

# Get the codelist for AVALCA1N when PARAMCD equals a specific condition
avalca1n_duration <- get_control_term(
  adex_spec,
  "AVALCA1N",
  where = "PARAMCD EQ ADURD"
)
avalca1n_duration
```

The `where` parameter accepts conditions that match the value-level
metadata definition from your specification. This information is mapped
to the `value_spec` table of your metacore object:

```{r eval=TRUE}
# See available VLM conditions
adex_spec$value_spec |>
  filter(variable == "AVALCA1N") |>
  pull(where)
```

#### **Handling Multiple VLM Conditions**

If you try to retrieve a VLM variable without specifying
a `where` condition, `get_control_term()` will alert you to the
available options:

```{r eval=TRUE, error=TRUE}
# No `where` provided
avalca1n_duration <- get_control_term(adex_spec, "AVALCA1N")
```

To see all possible codelists for a VLM variable, use `where = "all"`:

```{r eval=TRUE}
# Retrieve all codelists for this variable
all_avalca1n_codelists <- get_control_term(adex_spec, "AVALCA1N", where = "all")

# Result is a named list with unique names for each condition
names(all_avalca1n_codelists)
```

Each element in the returned list corresponds to one VLM condition and
contains its codelist:

```{r eval=TRUE}
# Access individual codelists by name
print(all_avalca1n_codelists$`ADEX: PARAMCD EQ ADURD`)
```

## Error Handling and Troubleshooting

#### **Error:** Variable not found in `value_spec`

```{r eval=TRUE, error=TRUE}
get_control_term(metacore, "NONEXISTENT_VAR")
```

**Solution:** Check the variable name against your specification.
Variable names are case-sensitive.

#### **Error:** Variable does not have a unique codelist. Please specify a dataset

```{r eval=TRUE, error=TRUE}
get_control_term(metacore, "QVAL")
```

**Solution:** Specify which dataset's version of the codelist you want

```{r eval=TRUE}
get_control_term(metacore, "QVAL", dataset = "SUPPAE")
```

#### **Error:** Variable has no controlled terminology

```{r eval=TRUE, error=TRUE}
get_control_term(metacore, "STUDYID")
```

**Solution:** Some variables may not have controlled terms (e.g.,
free-text fields). Confirm with your specification that this is expected
for your variable.

#### **Error:** No VLM condition found matching the supplied where clause

```{r eval=TRUE, error=TRUE}
get_control_term(adex_spec, "AVALCA1N", where = "INVALID WHERE")
```

**Solution:** Check the `metacore$value_spec` for your desired variable
to see a list of possible `where` conditions

```{r eval=TRUE}
adex_spec$value_spec |>
  filter(variable == "AVALCA1N") |>
  pull(where)
```

## Summary

The `get_control_term()` function provides a powerful, flexible way to
access standardized controlled terminology in your R workflows. By
leveraging metacore's metadata structure alongside companion packages
like metatools and xportr, you can:

-   Ensure data quality and consistency through validation against
    standardized codelists

-   Maintain regulatory compliance with traceable, reproducible
    workflows

-   Create reproducible, code that bridges raw data to regulatory
    submission

-   Handle both simple and complex (VLM) scenarios efficiently

-   Leverage the broader pharmaverse ecosystem for comprehensive
    metadata management

Whether you're validating data, creating derived variables, building
reference documentation, or preparing datasets for regulatory
submission, metacore puts standardized metadata at your fingertips.

For more information about metacore's structure and capabilities, see
the other vignettes and the [project
README](https://github.com/atorus-research/metacore). To learn more
about the broader ecosystem:

-   [metatools documentation](https://pharmaverse.github.io/metatools/)

-   [xportr documentation](https://atorus-research.github.io/xportr/)

-   [pharmaverse homepage](https://pharmaverse.org/)
